---
title: "PRISMA-Style Comparator Selection Flow Diagram"
subtitle: "EEE thesis candidates from classification through OpenAlex retrieval to comparator publication pool — LDP quasi-experimental study"
author: "Jason Pither"
date: today
format:
  html:
    self-contained: true
    toc: false
execute:
  echo: false
  warning: false
  message: false
---

## Purpose

This document renders a PRISMA-style flow diagram tracing EEE-classified thesis
candidates from the thesis classification project through eligibility filtering,
LDP student exclusion, and OpenAlex publication retrieval to the final comparator
publication pool used in the LDP FAIR compliance analysis.

### Data sources for all counts

| Count | Source |
|---|---|
| EEE theses (input) | `data/processed_data/classified/` CSVs from `LDP_thesis_classification` project |
| Filtering to eligible candidates | Computed from classified CSVs: `Category == "EEE"`, `prob_EEE ≥ 0.75`, `year ∈ 2022–2024`, 4 LDP target institutions |
| LDP student exclusion | `data/processed_data/private/ldp_exclusion_names.csv` (146 names; 28 overlapped with candidate pool) |
| OpenAlex retrieval | `scripts/02_get_comparator_authors.R`; counts at intermediate search steps were printed to console only |
| Final comparator pool | `data/processed_data/comparator_author_publications.csv` |

### Step descriptions

**EEE thesis classification.**
Thesis metadata from 8 Canadian institutions was classified by the
`LDP_thesis_classification` project pipeline. Of 16,900 clean theses, 1,682 were
predicted as EEE (ecology, evolution, or environment) by the trained text classifier.

**Eligibility filtering (`02_get_comparator_authors.R`).**
EEE-classified theses are restricted to the 4 institutions with LDP graduates
(McGill, UBC, Guelph, Manitoba), thesis submission years 2022–2024 (the years
covered by the scraped thesis data), and classifier probability ≥ 0.75. Within-
institution duplicate author entries are removed.

**LDP student exclusion.**
Candidates whose names appear on the LDP enrolment exclusion list
(`ldp_exclusion_names.csv`, 146 names) are removed to ensure no LDP participant
appears in the comparator pool.

**OpenAlex retrieval (two-phase search).**
For each of the 4 target institutions, a Phase 1 institution-level batch query
retrieves EEE-field works published from 2021 onwards and extracts unique first
authors. Candidate names are matched against the eligible pool using a first-initial +
last-name key; ambiguous matches are resolved by highest topic-discipline score.
Authors with more than 30 total OpenAlex works are excluded (likely non-students).
Phase 2 fetches the complete first-author publication list for each confirmed match.
Authors whose entire output matches non-EEE keyword patterns are not counted toward
the target. Comparators are collected at 2× the institution target N to ensure
sufficient coverage of publication years for downstream year-matched pairing.

---

```{r}
#| label: prisma-diagram
#| cache: false
#| fig-cap: >
#|   PRISMA-style flow of EEE thesis candidates from classification through
#|   eligibility filtering and OpenAlex retrieval to the final comparator
#|   publication pool. Blue boxes = main data flow; light yellow boxes =
#|   processing / search steps; dark green box = final output;
#|   red boxes = excluded records.

library(DiagrammeR)

dot_src <- '
digraph prisma {

  graph [rankdir  = TB,
         splines  = ortho,
         nodesep  = 0.55,
         ranksep  = 0.60,
         fontname = "Helvetica",
         bgcolor  = "white"]

  node [fontname = "Helvetica",
        fontsize  = 9,
        style     = "filled,rounded",
        shape     = rectangle,
        margin    = "0.18,0.12",
        penwidth  = 1.2]

  edge [arrowsize = 0.7,
        penwidth  = 1.0]

  //=================================================================
  // INPUT
  //=================================================================

  input [label    = "EEE-CLASSIFIED THESES\nOutput of LDP_thesis_classification project\nn = 1,682",
         fillcolor = "#D6EAF8", color = "#1A5276", width = 4.2,
         fontname  = "Helvetica-Bold"]

  //=================================================================
  // ELIGIBILITY FILTERING
  //=================================================================

  filter_step [label = "Eligibility filtering  (02_get_comparator_authors.R)\nRestrict to 4 LDP target institutions\n(McGill, UBC, Guelph, Manitoba)\nThesis years 2022–2024  |  prob_EEE ≥ 0.75\nRemove within-institution duplicate authors",
               fillcolor = "#FEF9E7", color = "#7D6608", width = 4.2,
               style = "filled,rounded"]

  eligible [label    = "ELIGIBLE CANDIDATES\n4 institutions\nn = 421 unique authors\n(McGill: 80  |  UBC: 188  |  Guelph: 93  |  Manitoba: 60)",
            fillcolor = "#D6EAF8", color = "#1A5276", width = 4.2]

  e_filter [label = "Excluded at eligibility filter\nn = 1,261\n(wrong institution, thesis year,\nor classifier confidence < 0.75)",
            fillcolor = "#FADBD8", color = "#922B21", width = 3.0]

  //=================================================================
  // LDP STUDENT EXCLUSION
  //=================================================================

  ldp_excl_step [label = "LDP student exclusion\nRemove candidates on LDP enrolment\nexclusion list  (n = 146 names on list;\nn = 28 overlapped with candidate pool)",
                 fillcolor = "#FEF9E7", color = "#7D6608", width = 4.2,
                 style = "filled,rounded"]

  candidates [label    = "ELIGIBLE NON-LDP CANDIDATES\n4 institutions\nn = 393 authors\n(McGill: 69  |  UBC: 171  |  Guelph: 93  |  Manitoba: 60)",
              fillcolor = "#D6EAF8", color = "#1A5276", width = 4.2]

  e_ldp [label = "Excluded: LDP students\nn = 28",
         fillcolor = "#FADBD8", color = "#922B21", width = 3.0]

  //=================================================================
  // OPENALEX SEARCH
  //=================================================================

  search_step [label = "OpenAlex publication retrieval  (02_get_comparator_authors.R)\nPhase 1: institution-level batch query (EEE field, from 2021)\nPhase 2: per-author targeted works fetch\nExclude authors with > 30 total works\nExclude publications matching non-EEE keyword patterns\nOversample at 2× institution target N for year-matched pairing\nTarget N: McGill = 7  |  UBC = 13  |  Guelph = 1  |  Manitoba = 1",
               fillcolor = "#FEF9E7", color = "#7D6608", width = 4.2,
               style = "filled,rounded"]

  //=================================================================
  // FINAL OUTPUT
  //=================================================================

  final [label = "COMPARATOR PUBLICATION POOL\n109 authors  |  262 publications",
         fillcolor = "#82E0AA", color = "#1D6A39", width = 4.2,
         style = "filled,bold,rounded", fontsize = 11]

  //=================================================================
  // MAIN FLOW EDGES
  //=================================================================

  input        -> filter_step
  filter_step  -> eligible
  eligible     -> ldp_excl_step
  ldp_excl_step -> candidates
  candidates   -> search_step
  search_step  -> final

  //=================================================================
  // EXCLUSION EDGES  (constraint = false: no rank effect)
  //=================================================================

  filter_step   -> e_filter [constraint = false, style = dashed, color = "#A93226"]
  ldp_excl_step -> e_ldp    [constraint = false, style = dashed, color = "#A93226"]

  //=================================================================
  // RANK GROUPS
  //=================================================================

  { rank = same;  eligible;   e_filter }
  { rank = same;  candidates; e_ldp    }

  //=================================================================
  // INVISIBLE ORDERING EDGES
  //=================================================================

  eligible   -> e_filter [style = invis]
  candidates -> e_ldp    [style = invis]

}
'

grViz(dot_src, height = 800, width = "100%")
```

---

```{r}
#| label: export-svg-pdf
#| cache: false

library(DiagrammeRsvg)
library(rsvg)
library(here)

svg_str  <- DiagrammeRsvg::export_svg(grViz(dot_src))
svg_path <- here::here("scripts", "comparator_selection_prisma.svg")
pdf_path <- here::here("scripts", "comparator_selection_prisma.pdf")

writeLines(svg_str, svg_path)
rsvg::rsvg_pdf(svg = svg_path, file = pdf_path)

message("Exported:\n  ", svg_path, "\n  ", pdf_path)
```
